---
title: "Shiny Application Pitch"
author: "Konstantinos"
date: "25 December 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Intro 

- I used the Iris dataset to demonstrate a simple idea
```{r}
summary(iris)
```
## The problem

- We have 4 variable that we want to use in order to classify the 5th variable Species or to predict Species (setosa, versicolor, virginica )
- It is difficult to find out if these 4 variables help us 'seperate' the three Species. Imagine if we had even more variables.
- **My Solution:** *Principal Components Analysis*(**PCA**). Dimensionality Reduction method. I will visualise the separation of the three species with the Principal Components.

## What the app does? How is it usefull?
- It is just a simple, very basic app. You can only choose how many Principal Components the scatter plot matrix has.
- You can see that how many PCs you need to separate the Species. You can use the PCs in furute analysis or in a predictive model instead of the original 4 variables. 
- As you can see bellow the first two commponents explain 95% of the variability of the dataset.
```{r}
pca <- prcomp(iris[,-5],
                center = TRUE,
                scale = TRUE)
summary(pca)
```
## Where to find it?
- https://konanthis.shinyapps.io/Visualize_Multidimensional_Data/
- https://github.com/AnthisKon/datasciencecoursera/

- Thank you!
